Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-07-19 tex/manuscript.tex
Update manuscript.tex

% List of authors, with corresponding author marked by asterisk
\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[8pt]
EIVIND HOVIG$^\ast$\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[2pt]
\textit{\NOTE{Andre adresser}}
\\[2pt]
% E-mail address for correspondence
{ehovig@ifi.uio.no}}

@@

\begin{abstract}
{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous. It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments. Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner. The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
}
\end{abstract}

@@

Extraneous variables, if left unaccounted for, have the potential to lead an investigator into drawing wrong conclusions. A common example is ``batch effects'' caused by reagents, microarray chips, and other equipment made in batches, which often have systematic effects on the measurements. A similar example is ``centre effects'' when samples or data come from multiple sources. See \citet{Luo2010} for more examples.

For a typical experiment comparing differences between study groups, the presence of batch effects will decrease the statistical power since it adds variation to the data. If the batch--group design is unbalanced, i.e if the study groups are not equally represented in all batches, batch effects may also act as a confounder and induce false differences between groups (\citealp{Leek2010}).

The standard way to handle an extraneous variable is to include it in the statistical model employed in the inquiry. However, many analysis tools for high throughput data do not cater for this option, and when available it could still be outside the competence of the investigator. Therefore, an alternative two step procedure has emerged. First the batch effects are estimated and removed, creating a ``batch effect free'' data set. Then, the statistical analyses are performed on the adjusted data without further consideration of batch effects. This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel. Unfortunately, as we demonstrate in this paper, when the batch--group design is unbalanced, this approach may be unreliable.

A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R. When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power. However, when the batch--group design is unbalanced, batch differences will in part be influenced by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power. In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way. Figure~\ref{fig:boxplots} illustrates both these effects.

@@

The ComBat method described in \citet{Johnson2007}, and included in the \texttt{sva} package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences. In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora, and ComBat has thus help popularise this approach. Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analyses. And as a consequence, confidence in group effects has been overestimated and false results reported.

Statisticians would most likely take extra precautions should the sample or batch sizes be very small. However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain even as sample and batch sizes increase. For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design rather than the size of the sample or batches. The effect of this may be further exacerbated by running these analyses a large number of times, e.g. on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.

>>>>

% List of authors, with corresponding author marked by asterisk
\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[8pt]
EIVIND HOVIG$^\ast$\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[2pt]
\textit{Institute of Cancer Genetics and Informatics, Oslo University Hospital HF - Radiumhospitalet,
Montebello, 
0310 Oslo,
Norway}
\textit{Department of Informatics, University of Oslo, 
0316 OSLO
Norway
Norway}
\\[2pt]
% E-mail address for correspondence
{ehovig@ifi.uio.no}}

@@

\begin{abstract}
{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat, which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous. It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments. Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner. The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
}
\end{abstract}

@@

Extraneous variables, if left unaccounted for, have the potential to lead an investigator into drawing wrong conclusions. A common example is ``batch effects'' caused by reagents, microarray chips, and other equipment made in batches that may vary in some way, which often have systematic effects on the measurements. A similar example is ``centre effects'', when samples or data come from multiple sources. See \citet{Luo2010} for more examples.

For a typical experiment of comparing differences between study groups, the presence of batch effects will decrease the statistical power, since it adds variation to the data. If the batch--group design is unbalanced, i.e. if the study groups are not equally represented in all batches, batch effects may also act as a confounder and induce false differences between groups (\citealp{Leek2010}).

The standard way to handle an extraneous variable is to include it in the statistical model employed in the inquiry. However, many analysis tools for high throughput data do not cater for this option, and when available, it could still be outside the competence of the investigator. Therefore, an alternative two step procedure has emerged. First the batch effects are estimated and removed, creating a ``batch effect free'' data set. Then, the statistical analyses are performed on the adjusted data without further consideration of batch effects. This appealing compartmentalization is also convenient for practical purposes, for example when data-processing and statistical analyses are performed by different personnel. Unfortunately, as we demonstrate in this paper, when the batch--group design is unbalanced, this approach may be unreliable.

A simple removal of batch effects can be achieved by subtracting the mean of the measurements in one batch from all measurements in that batch, i.e mean adjustment or one-way ANOVA adjustment as implemented in the method \texttt{pamr.batchadjust} from the \texttt{pamr} package in R. When the batch--group design is balanced, mean-adjustment will remove most, but not necessarily all, variance attributed to batch and leave the between group variance, thus increasing the statistical power. However, when the batch--group design is unbalanced, batch differences will in part be influenced by group differences, and thus batch correction will reduce group differences and thereby reduce the statistical power. In very uneven group--batch designs with multiple groups, spurious group differences may even be induced in this way. Figure~\ref{fig:boxplots} illustrates both of these effects.

@@

The ComBat method described in \citet{Johnson2007}, and included in the \texttt{sva} package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences. In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora, and ComBat has thus helped popularise this approach. Based on the actual use of ComBat by the authors and others, ourselves included, we suspect thus that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses. And as a consequence, confidence in group effects has been overestimated and false results reported.

Statisticians would most likely take extra precautions, should the sample or batch sizes be very small. However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain even as sample and batch sizes increase. For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches. The effect of this may be further exacerbated by running these analyses a large number of times, e.g. on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.
